---
title: "R Notebook"
output: html_notebook
---

# Purpose
This is a notebook to compare the different deweathering approaches to compute y-o-y changes in emissions.

We'll look at the following models:
- trend, with various time variables
- anomaly, after excluding the months of interest from training dataset
- anomaly, without excluding the months of interest from training dataset

```{r, include=FALSE, echo=FALSE}

library(tidyverse)
library(creadeweather)
readRenviron(".Renviron")

# Use current url to collect NCAP locations
anomalies <- read_csv("https://api.energyandcleanair.org/ncap/anomaly?pollutant=pm25&ncap_only=true&year=2024&month=8&deweather_method=default_anomaly_2018_2099,&format=csv")
locations_ncap <- unique(anomalies$location_id)
locations_mee <- read_csv("https://api.energyandcleanair.org/cities?country=CN&format=csv")$id

# take 10 random locations
# set seed
set.seed(123)
locations <- c(
  sample(locations_ncap, 1),
  sample(locations_mee, 1))
weather_file <- "tmp/weather_yoy.RDS"
# file.remove(weather_file)
month_yoy <- "2024-07-01"

```


```{r, cache=TRUE}
deweathered_trend <- creadeweather::deweather(
  location_id = locations,
  poll = c("pm25", "no2"),
  source = c("cpcb", "mee"),
  tree = 20000,
  learning.rate = 0.01,
  interaction.depth = 7,
  training.fraction = 1,
  lag = 1,
  weather_vars = c("air_temp_min","air_temp_max","atmos_pres","wd","ws","precip","dewpoint_temp","pbl_min","pbl_max"),
  time_vars = c("date_unix"),
  upload_results = F,
  save_weather_filename = weather_file,
  read_weather_filename = weather_file,
  weather_update_era5 = F
)


deweathered_trend_2 <- creadeweather::deweather(
  location_id = locations,
  tree = 20000,
  learning.rate = 0.01,
  interaction.depth = 7,
  training.fraction = 1,
  lag = 1,
  weather_vars = c("air_temp_min","air_temp_max","atmos_pres","wd","ws","precip","dewpoint_temp","pbl_min","pbl_max"),
  time_vars = c("date_unix", "yday"),
  poll = c("pm25", "no2"),
  source = c("cpcb", "mee"),
  upload_results = F,
  save_weather_filename = weather_file,
  read_weather_filename = weather_file,
  weather_update_era5 = F
)


deweathered_trend_3 <- creadeweather::deweather(
  location_id = locations,
  tree = 20000,
  learning.rate = 0.01,
  interaction.depth = 7,
  training.fraction = 1,
  lag = 1,
  weather_vars = c("air_temp_min","air_temp_max","atmos_pres","wd","ws","precip","dewpoint_temp","pbl_min","pbl_max"),
  time_vars = c("date_unix", "yday", "wday"),
  poll = c("pm25", "no2"),
  source = c("cpcb", "mee"),
  upload_results = F,
  save_weather_filename = weather_file,
  read_weather_filename = weather_file,
  weather_update_era5 = F
)
```

Let's plot time series
```{r}
bind_rows(
  deweathered_trend %>% mutate(time_vars = "date_unix"),
  deweathered_trend_2 %>% mutate(time_vars = "date_unix + yday"),
  deweathered_trend_3 %>% mutate(time_vars = "date_unix + yday + wday")
) %>%
  select(location_id, source, poll, time_vars, result) %>%
  tidyr::unnest(result) %>%
  filter(variable == "trend") %>%
  ggplot(aes(date, value, color = time_vars)) +
  geom_line() +
  facet_wrap(~location_id + poll, scales = "free_y") +
  theme_minimal() +
  theme(legend.position = "bottom")

```

# Comparing YOYs with anomaly approach

```{r}
deweathered_anomaly <- creadeweather::deweather(
  location_id = locations,
  deweather_process_id = "default_anomaly_2018_2099",
  time_vars = c(),
  poll = c("pm25", "no2"),
  source = c("cpcb", "mee"),
  upload_results = F,
  save_weather_filename = weather_file,
  read_weather_filename = weather_file,
  weather_update_era5 = F
)
```

```{r}
extract_yoys_from_trend <- function(deweathered_trend) {
  deweathered_trend %>%
    select(location_id, source, poll, models, result) %>%
    tidyr::unnest(result) %>%
    group_by(location_id,
             source,
             poll,
             models,
             year = year(date),
             month = month(date),
             variable) %>%
    summarise(value = mean(value, na.rm = T)) %>%
    arrange(year) %>%
    group_by(across(-c(year, value))) %>%
    mutate(delta = value - lag(value)) %>%
    # Add observed value of first year
    ungroup() %>%
    group_by(location_id, source, poll, models, month, year) %>%
    mutate(observed_prev = value[variable == "observed"] - delta[variable == "observed"]) %>%
    ungroup() %>%
    filter(!is.na(delta)) %>%
    select(-c(value)) %>%
    tidyr::pivot_wider(
      names_from = "variable",
      values_from = "delta",
      names_prefix = "yoy_"
    ) %>%
    rename(
      yoy_total = yoy_observed,
      yoy_emission = yoy_trend,
    ) %>%
    mutate(
      yoy_weather = yoy_total - yoy_emission,
      yoy_total_rel = yoy_total / observed_prev,
      yoy_residual = yoy_anomaly / observed_prev,
      yoy_weather_rel = yoy_weather / observed_prev,
      yoy_emission_rel = yoy_emission / observed_prev
    ) %>%
    pivot_longer(
      cols = -c(year, observed_prev, location_id, source, models, poll),
      names_to = "variable",
      values_to = "value",
    ) %>%
    filter(grepl("yoy.*_rel", variable)) %>%
    select(location_id, source, poll, models, variable, value) %>%
    mutate(date=month_yoy) %>%
    group_by(location_id, source, poll, models) %>%
    tidyr::nest() %>%
    rename(result = data)
}

```


```{r}
yoys <- bind_rows(
  extract_yoys_from_trend(deweathered_trend) %>% mutate(model = "trend"),
  extract_yoys_from_trend(deweathered_trend_2) %>% mutate(model = "trend_2"),
  extract_yoys_from_trend(deweathered_trend_3) %>% mutate(model = "trend_3")
) %>%
  unnest(result) 
```


Let's compare Yoys

```{r}
yoys %>%
  group_by(location_id, source, model) %>%
  summarise(
    yoy_total = mean(value[variable == "yoy_total"]),
    yoy_emission = mean(value[variable == "yoy_emission"]),
    yoy_weather = mean(value[variable == "yoy_weather"]),
    yoy_total_rel = mean(value[variable == "yoy_total_rel"]),
    yoy_emission_rel = mean(value[variable == "yoy_emission_rel"]),
    yoy_weather_rel = mean(value[variable == "yoy_weather_rel"]),
    n = n()
  ) %>%
  ungroup() %>%
  arrange(location_id, source, model)
